---
title: "Linear mixed effects models"
author: "Aleksander Wael"
date: "9/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(lme4)
library(tidyverse, lme4)
```

# Tutorial Part 1: Linear modeling

# Modelling pitch and sex

```{r}
pitch = c(233,204,242,130,112,142)
sex = c(rep("female",3),rep("male",3))
```

# Make the df

```{r}
my.df <- tibble(sex, pitch)
head(my.df)
```

# Make linear model

```{r}
xmdl = lm(pitch ~ sex, my.df)
summary(xmdl)
```
# Checking that intercept is in fact the mean because there is only two categories

```{r}
mean(my.df[my.df$sex=="female",]$pitch)
```
# Doing same thing but with continuos data

```{r}
age = c(14,23,35,48,52,67)
pitch = c(252,244,240,233,212,204)
my.df = tibble(age,pitch)
xmdl = lm(pitch ~ age, my.df)
summary(xmdl)
```

# Intercept isn't meaningful (you don't have a voice pitch at age 0). Centering variable around mean.

```{r}
my.df$age.c = my.df$age - mean(my.df$age)
xmdl = lm(pitch ~ age.c, my.df)
summary(xmdl)
```

# Making residual plot

```{r}
plot(fitted(xmdl),residuals(xmdl))
```

# Checking assumptions
# (1) Linearity (that the two variables are in fact linear, check visually and with residuals)
# (2) Absence of collinearity (dont know how to check, maybe visually?)
# (3) Homoskedasticity (same variance of residuals, please)
# (4) Normality of residuals (histogram of residuals)
# (5) Absence of influential data points (leave-one-out, how influential is each data point for my model)
# (6) Independence !!!!!!! (A coin flip is independent from the result of the next coin flip, no learning effects etc.)


```{r}
# Using par() to show both plots, just to check if the qq norm is same in both plots
par(plot(xmdl), qqnorm(residuals(xmdl)), hist(residuals(xmdl)))
```

# Checking absence of influential data points
# This is actually easy leave-one-out!

```{r}
dfbeta(xmdl)
```

# Part 2: A very basic tutorial for performing linear mixed effects analyses

```{r}
# Load data
politeness <- read.csv("http://www.bodowinter.com/tutorial/politeness_data.csv")
view(politeness)
```

```{r}
# See NA (dont have to remove for mixed effects model)
which(is.na(politeness)==T)
```

```{r}
# Boxplot to see differences by varbiables
boxplot(frequency ~ attitude*gender, col=c("white","lightgray"),politeness)
```

```{r}
# Modelling with mixed and random effects
# REML = FALSE gives info about model explanation power
politeness.model <- lmer(frequency ~ attitude + gender + (1|subject) + (1|scenario), data=politeness, REML = FALSE)
summary(politeness.model)

# Making another model (one without attitude) to see if my fixed effects differ
politeness.null <- lmer(frequency ~ gender + (1|subject) + (1|scenario), data=politeness, REML=FALSE)

# Comparing the models
anova(politeness.null,politeness.model)

# We see a significant difference os the models, so the attitude has an effect
```

```{r}
# Modelling random slopes
politeness.model = lmer(frequency ~ attitude + gender + (1+attitude|subject) + (1+attitude|scenario), data=politeness, REML=FALSE)

coef(politeness.model)
```

