---
title: "practical_exercise_3, Methods 3, 2021, autumn semester"
author: "Aleksander Moeslund Wael"
date: "01/10/2021"
output:
  pdf_document: default
  html_document: default
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages

```{r}
pacman::p_load(tidyverse, lmerTest, lme4)
```

# Exercise 1

## 1. Creating a data frame with all subject data

```{r}
files <- list.files(path = "experiment_2",         
                    pattern = ".csv",
                    full.names = T)

df <- read_csv(files)
```
  
## 2. Describing the data
  
  The dataset contains 18131 observations described by 17 variables. Data from 29 subjects is included. 

### 2. i.
  
  Adding variable "correct" to display if subject was correct
  
```{r}
# Adding empty variable
df <- df %>% 
  mutate(obj.resp.2 = obj.resp)

# Renaming rows in obj.resp.2 to get same units as target.type
df$obj.resp.2 <- replace(df$obj.resp.2, df$obj.resp.2 == "e", "even")
df$obj.resp.2 <- replace(df$obj.resp.2, df$obj.resp.2 == "o", "odd")

# Adding value for correct and incorrect answers
df_correct <- df %>%
  filter(obj.resp.2 == target.type) %>% 
  mutate(correct = "1")

# Joining with my df
df <- left_join(df, df_correct)

# Remaining are NAs, so replace with 0
df$correct <- replace(df$correct, is.na(df$correct), "0")
```

### 2. ii. Describe what the following variables in the data frame contain, trial.type, pas, trial, target.contrast, cue, task, target_type, rt.subj, rt.obj, obj.resp, subject and correct. (That means you can ignore the rest of the variables in your description). For each of them, indicate and argue for what class they should be classified into, e.g. factor, numeric etc.

```{r, EVAL = FALSE}
unique(df$trial.type)
unique(df$pas)
unique(df$trial)
unique(df$target.contrast)
unique(df$cue)
test <- df %>% 
  filter(subject == "001", cue == "0")
unique(test$cue)
unique(df$task)
unique(df$target.type)
```

trial.type: Indicates whether subject is doing the staircase task (first experiment) or the follow-up experiment.
Should be class character, as it is a category.
pas: Indicates subjects response to trial on the Perceptual Awareness Scale (PAS). Takes a value between 1-4, and will therefore be treated as numeric.
trial: A numbered list for every trial the subject completes, i.e. presses e or o in either of the trial types., per subject. I should think character class for now (might change).
target.contrast: The contrast between the background and the digit (target). Between 0-1, treated as numeric.
cue: The specific cue pattern, will treat as character.
task: Whether cue pattern is 2 (singles), 4 (pairs) or 8 (quadruplets) digits. Will treat as character.
target.type: Whether target type is an odd or even number - will treat as character.
rt.subj: Reaction time for response to PAS pr. trail - will treat as numeric.
rt.obj: Reaction time for responding if target is even or odd - will treat as numeric.
obj.resp: Subjects response to target is either even or odd - will treat as character.
subject: Participant ID, ordered from 001. Treated as character.
correct: Whether subject answered correctly in the trail, 1 for correct and 0 for incorrect. Is logical (binary)

```{r}
# Assigning variables to proper class
df$pas <- as.numeric(df$pas)
df$trial <- as.character(df$trial)
df$target.contrast <- as.numeric(df$target.contrast)
df$cue <- as.character(df$cue)
df$rt.subj <- as.numeric(df$rt.subj)
df$rt.obj <- as.numeric(df$rt.obj)
df$target.contrast <- as.numeric(df$target.contrast)
df$correct <- as.integer(df$correct)
df$correct <- as.logical(df$correct)
```

### 2. iii. for the staircasing part only, create a plot for each subject where you plot the estimated function (on the target.contrast range from 0-1) based on the fitted values of a model (use glm) that models correct as dependent on target.contrast. These plots will be our no-pooling model. Comment on the fits - do we have enough data to plot the logistic functions?

```{r}
# Making df
df_staircase <- df %>% 
  filter(trial.type == "staircase")

# Building model (complete pooling)
m1 <- glm(correct ~ target.contrast, data = df_staircase, family = "binomial")

# Plotting
df_staircase %>% 
  ggplot(aes(x = target.contrast, y = fitted(m1), color = correct)) +
  geom_point() +
  facet_wrap(~subject)+
  theme_bw()
```

### 2.iv. on top of those plots, add the estimated functions (on the target.contrast range from 0-1) for each subject based on partial pooling model (use glmer from the package lme4) where unique intercepts and slopes for target.contrast are modelled for each subject

```{r}
# Building model (partial-pooling)
m2 <- glmer(correct ~ target.contrast + (1 + target.contrast|subject) , data = df_staircase, family = "binomial")
effects <- ranef(m2)
effects <- effects$subject
effects <- rownames_to_column(effects, "subject")
m2_intercepts <- effects$subject[1]
m2_slopes <- effects$subject[2]

# Plotting
df_staircase %>% 
  ggplot(aes(x = target.contrast, y = fitted(m1), color = correct))+
  geom_point()+
  geom_line(aes(x = target.contrast, y = fitted(m2), color = "black"))+
  facet_wrap(~subject)+
  theme_bw()

```

### 2. v. in your own words, describe how the partial pooling model allows for a better fit for each subject

Partial pooling allows for the model to be generalizable, but still accounts for subject differences in baseline (intercept) and performance (slopes).

# Exercise 2

```{r}
# Making df
df_experiment <- df %>% 
  filter(trial.type == "experiment")
```

### 2. 1. Pick four subjects and plot their Quantile-Quantile (Q-Q) plots for the residuals of their objective response times (rt.obj) based on a model where only intercept is modelled

```{r}
df_experiment <- df %>% 
  filter(trial.type == "experiment")
response_time <- lm(rt.obj ~ 1, data = df_experiment)
df_experiment$fitted_rt <- fitted(response_time)
```

```{r}
pacman::p_load(car)
subject1 <- df_experiment %>% 
  filter(subject == "001")
subject2 <- df_experiment %>% 
  filter(subject == "002")
subject3 <- df_experiment %>% 
  filter(subject == "003")
subject4 <- df_experiment %>% 
  filter(subject == "004")
interceptmodel1 <- lm(rt.obj ~ 1, data = subject1)
interceptmodel2 <- lm(rt.obj ~ 1, data = subject2)
interceptmodel3 <- lm(rt.obj ~ 1, data = subject3)
interceptmodel4 <- lm(rt.obj ~ 1, data = subject4)
qqPlot(interceptmodel1)
qqPlot(interceptmodel2)
qqPlot(interceptmodel3)
qqPlot(interceptmodel4)
```

### 2. 1. i. comment on these


### 2. 1. ii. does a log-transformation of the response time data improve the Q-Q-plots?
```{r}
logsubject1 <- subject1 %>% 
  mutate(log_rt = log(rt.obj))
logsubject2 <- subject2 %>% 
  mutate(log_rt = log(rt.obj))
logsubject3 <- subject3 %>% 
  mutate(log_rt = log(rt.obj))
logsubject4 <- subject4 %>% 
  mutate(log_rt = log(rt.obj))
loginterceptmodel1 <- lm(log_rt ~ 1, data = logsubject1)
loginterceptmodel2 <- lm(log_rt ~ 1, data = logsubject2)
loginterceptmodel3 <- lm(log_rt ~ 1, data = logsubject3)
loginterceptmodel4 <- lm(log_rt ~ 1, data = logsubject4)
qqPlot(loginterceptmodel1)
qqPlot(loginterceptmodel2)
qqPlot(loginterceptmodel3)
qqPlot(loginterceptmodel4)
```


### 2. 2. Now do a partial pooling model modelling objective response times as dependent on _task_? (set `REML=FALSE` in your `lmer`-specification) 
```{r}
rt_partialpooling1 <- lmer(rt.obj ~ task + (1|subject), REML = FALSE, data = df_experiment)
rt_partialpooling2 <- lmer(rt.obj ~ task + (1|subject) + (1|trial), REML = FALSE, data = df_experiment)
```

### 2. 2. i. which would you include among your random effects and why? (support your choices with relevant measures, taking into account variance explained and number of parameters going into the modelling)  


### 2. 2. ii. explain in your own words what your chosen models says about response times between the different tasks  


### 2. 3. Now add _pas_ and its interaction with _task_ to the fixed effects 
```{r}
rt_partialpooling3 <- lmer(rt.obj ~ task*pas + (1|subject), REML = FALSE, data = df_experiment)
summary(rt_partialpooling3)
```

### 2. 3. i. how many types of group intercepts (random effects) can you add without ending up with convergence issues or singular fits?
```{r}
rt_partialpooling4 <- lmer(rt.obj ~ task*pas + (1|subject) + (1|trial), REML = FALSE, data = df_experiment)
rt_partialpooling5 <- lmer(rt.obj ~ task*pas + (1|subject) + (1|trial) + (1|odd.digit), REML = FALSE, data = df_experiment)
rt_partialpooling6 <- lmer(rt.obj ~ task*pas + (1|subject) + (1|trial) + (1|odd.digit) + (1|cue), REML = FALSE, data = df_experiment)
rt_partialpooling7 <- lmer(rt.obj ~ task*pas + (1|subject) + (1|trial) + (1|odd.digit) + (1|cue) + (1|pas), REML = FALSE, data = df_experiment)
rt_partialpooling8 <- lmer(rt.obj ~ task*pas + (1|subject) + (1|trial) + (1|odd.digit) + (1|cue) + (1|pas) + (1|seed), REML = FALSE, data = df_experiment)
rt_partialpooling9 <- lmer(rt.obj ~ task*pas + (1|subject) + (1|trial) + (1|odd.digit) + (1|cue) + (1|pas) + (1|seed) + (1|even.digit), REML = FALSE, data = df_experiment) #Finally 
```


### 2. 3. ii. create a model by adding random intercepts (without modelling slopes) that results in a singular fit - then use `print(VarCorr(<your.model>), comp='Variance')` to inspect the variance vector - explain why the fit is singular (Hint: read the first paragraph under details in the help for `isSingular`)

```{r}
print(VarCorr(rt_partialpooling9), comp='Variance')
```

The fit is singular, because "even.digit" explains no variance. 

### 2. 3. iii. in your own words - how could you explain why your model would result in a singular fit?

"even.digit" explains no variance, because I had already added "odd.digit", and they explain the samevariance. 

# Exercise 3

```{r}
data.count <- df %>%
  group_by(subject, task, pas) %>%
  dplyr::summarise("count" = n())
```        

### 3. 2. Now fit a multilevel model that models a unique "slope" for _pas_ for each _subject_ with the interaction between _pas_ and _task_ and their main effects being modelled  

```{r}
pasmodel <- glmer(count ~ pas*task + (pas|subject), data = data.count, family = poisson, control = glmerControl(optimizer="bobyqa"))
summary(pasmodel)
```

### 3. 2. i. which family should be used?  
Poisson, because it's good for modelling counts. 

### 3. 2. ii. why is a slope for _pas_ not really being modelled?  

Because pas isn't continuous, we can't model a proper slope, so this is kind of a "pseudo"-slope.

### 3. 2. iii. if you get a convergence error, try another algorithm (the default is the _Nelder_Mead_) - try (_bobyqa_) for which the `dfoptim` package is needed. In `glmer`, you can add the following for the `control` argument: `glmerControl(optimizer="bobyqa")` (if you are interested, also have a look at the function `allFit`)

I did get the error, now it works :) 

### 3. 2. iv. when you have a converging fit - fit a model with only the main effects of _pas_ and _task_. Compare this with the model that also includes the interaction


### 3. 2. v. indicate which of the two models, you would choose and why  


### 3. 2. vi. based on your chosen model - write a short report on what this says about the distribution of ratings as dependent on _pas_ and _task_  

```{r}
pasmodel2 <- glmer(count ~ pas + task + (pas|subject), data = data.count, family = poisson)
tibble(sum(residuals(pasmodel)^2), sum(residuals(pasmodel2)^2))
AIC(pasmodel, pasmodel2)
```

### 3. 2. vii.

```{r}
pas_foursubjects <- data.count %>% 
  filter(subject == "001"|subject == "002"|subject == "003"|subject == "004")
pasmodel_foursubjects <- glmer(count ~ pas + task + (pas|subject), data = pas_foursubjects, family = poisson)
pas_foursubjects %>% 
  ggplot() +
  geom_point(aes(x = pas, y = fitted(pasmodel_foursubjects), color = "Estimated")) +
  geom_point(aes(x = pas, y = count, color = "Observed")) +
  facet_wrap( ~ subject)
```


### 3. 3. Finally, fit a multilevel model that models correct as dependent on task with a unique intercept for each subject

```{r}
df_end_me <- glmer(correct ~ task + (1 | subject), data = df, family = "binomial")
```

### 3. 3. i. does task explain performance?

```{r}
summary(df_end_me)
```

Task significantly predicts correctness for all task levels (all p < 0.05).

### 3. 3. ii. add pas as a main effect on top of task - what are the consequences of that?

```{r}
df_end_me_more <- glmer(correct ~ task + pas + (1 | subject), data = df, family = "binomial")
summary(df_end_me_more)
```

Since task is no longer significant, it seems that pas explains more of the variance, i.e. a better predictor.

### 3. 3. iii. now fit a multilevel model that models correct as dependent on pas with a unique intercept for each subject

```{r}
df_end_me_more_now <- glmer(correct ~ pas + (1 | subject), data = df, family = "binomial")
summary(df_end_me_more_now)
```

### 3. 3. iv. finally, fit a model that models the interaction between task and pas and their main effects

```{r}
df_end_me_more_now_pls <- glm(correct ~ task * pas, data = df, family = "binomial")
summary(df_end_me_more_now_pls)
```

### 3. 3. v. describe in your words which model is the best in explaining the variance in accuracy.

The model which predicts correct by pas with intercepts per subject has the lowest AIC value and a significant chi-square value. This model strikes the balance between complexity and explanatory power.

