---
title: "practical_exercise_3, Methods 3, 2021, autumn semester"
author: 'Aleksander Moeslund Wael'
date: "01/10/2021"
output: html_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages

```{r}
pacman::p_load(tidyverse)
```

# Exercise 1

## 1. Creating a data frame with all subject data

```{r}
files <- list.files(path = "experiment_2",         
                    pattern = ".csv",
                    full.names = T)

df <- read_csv(files)
```
  
## 2. Describing the data
  
  The dataset contains 18131 observations described by 17 variables. Data from 29 subjects is included. 

### 2. i.
  
  Adding variable "correct" to display if subject was correct
  
```{r}
# Adding empty variable
df <- df %>% 
  mutate(obj.resp.2 = obj.resp)

# Renaming rows in obj.resp.2 to get same units as target.type
df$obj.resp.2 <- replace(df$obj.resp.2, df$obj.resp.2 == "e", "even")
df$obj.resp.2 <- replace(df$obj.resp.2, df$obj.resp.2 == "o", "odd")

# Adding value for correct and incorrect answers
df_correct <- df %>%
  filter(obj.resp.2 == target.type) %>% 
  mutate(correct = "1")

# Joining with my df
df <- left_join(df, df_correct)

# Remaining are NAs, so replace with 0
df$correct <- replace(df$correct, is.na(df$correct), "0")
```

### 2. ii. Describe what the following variables in the data frame contain, trial.type, pas, trial, target.contrast, cue, task, target_type, rt.subj, rt.obj, obj.resp, subject and correct. (That means you can ignore the rest of the variables in your description). For each of them, indicate and argue for what class they should be classified into, e.g. factor, numeric etc.

```{r}
unique(df$trial.type)
unique(df$pas)
unique(df$trial)
unique(df$target.contrast)
unique(df$cue)
test <- df %>% 
  filter(subject == "001", cue == "0")
unique(test$cue)
unique(df$task)
unique(df$target.type)
```

trial.type: Indicates whether subject is doing the staircase task (first experiment) or the follow-up experiment
pas: Indicates subjects response to trial on the Perceptual Awareness Scale (PAS)
trial: A numbered list for every trial the subject completes, i.e. presses e or o in either of the trial types. Per subject.
target.contrast: The contrast between the background and the digit (target)
cue: The specific cue pattern
task: Whether cue pattern is 2,4 or 8 digits
target.type: Whether target type is an odd or even number
rt.subj: Reaction time for response to PAS pr. trail.
rt.obj: Reaction time for responding if target is even or odd
obj.resp: Subjects response to target is either even or odd
subject: Participant ID, ordered from 001.
correct: Whether subject answered correctly in the trail, 1 for correct and 0 for incorrect.
